Testing and Isolation Efficacy: Insights from a Simple Epidemic Model

### 2021-04-28 SIR_manuscript.tex
R0 function was modified: S_n scaled by eta_w, also expressions A B modified to match the manuscript
model updated with (1-theta_w)

\begin{figure}[h!]
\centering
\begin{subfigure}[t]{.45\textwidth}
\centering
\includegraphics[width=\linewidth]{codes/R0contour_random_rho.5.pdf}
\caption{}
\end{subfigure}
%
\begin{subfigure}[t]{.45\textwidth}
\centering
\includegraphics[width=\linewidth]{codes/R0contour_TTI_rho.5.pdf}
\caption{}
\end{subfigure}
\caption{
{\bf A comparison of the behaviour of the basic reproduction number, $\Rnum$, between random versus targeted testing strategies at different levels of testing and isolation.}
We numerically evaluate $\Delta$ \eqref{R0}, reflecting the reduction of $\Rnum$ with respect to testing and isolation. We use the following parameters (listed in Table \ref{tab:params}):
$N_0=1 \times 10^6$, $\omega \in (0.5,2]$ 1/day, $1/\gamma= 3$ days, $\rho \in [0,0.5]$ 1/day, $\theta\_w$ and $\theta\_c$ vary between 0 and 1 with 0 for no effect and 1 for full effect of isolation on the transmission probability, $\beta=0.39$ 1/day, $p_S=0$, $p_I=1$ and $p_R=0.5$. Contours of $\Delta$ are plotted for two testing strategies identified by a set of relative testing weights; (a) random testing where $w_S=w_I=w_R=1$ and (b) targeted testing where $w_S=0.3$ and $w_I=w_R=1$. The black solid line in each panel represents the critical contour of $\Delta=1-\frac{\gamma}{\beta}$, i.e., the $\Delta$ corresponding to the threshold of $\Rnum=1$. \bmb{we still have some collisions in the y-axis tick labels}
}
\label{pan2}
\end{figure}

@@

The model in the form of a system of ordinary differential equations is 
\begin{subequations}\label{model}
\begin{align}
 d S\_u/dt &= -\Lambda S\_u - F_S S\_u + \omega S\_n, \\
 d S\_n/dt &= -\Lambda S\_n + (1-p_S) F_S S\_u - \omega S\_n, \\
 d S\_p/dt &= -\Lambda S\_p + p_S F_S S\_u - \omega S\_p, \\
 d S\_c/dt &= -\Lambda S\_c + \omega S\_p, \\
 d I\_u/dt &= \Lambda S\_u - F_I I\_u + \omega I\_n  - \gamma I\_u, \\
 d I\_n/dt &= \Lambda S\_n + (1-p_I) F_I I\_u - \omega I\_n -\gamma I\_n, \\
 d I\_p/dt &= \Lambda S\_p + p_I F_I I\_u - \omega I\_p -\gamma I\_p, \\
 d I\_c/dt &= \Lambda S\_c + \omega I\_p - \gamma I\_c, \\
 d R\_u/dt &= \gamma I\_u - F_R R\_u + \omega R\_n, \\
 d R\_n/dt &= \gamma I\_n + (1-p_R) F_R R\_u - \omega R\_n, \\
 d R\_p/dt &= \gamma I\_p + p_R F_R R\_u  - \omega R\_p, \\
 d R\_c/dt&= \gamma I\_c + \omega R\_p, \\
 dN/dt &= \omega (S\_n + I\_n + R\_n),  \\
 dP/dt &= \omega(I\_p + R\_p) ,
\end{align}
\end{subequations}

where parameters are specified in Table \ref{tab:params}. The next generation matrix for this model is $G = F V^{-1}$, where matrix $F$ represents the inflow of new infection to the infected compartments and matrix $V$ represents the flow in the infected compartments when the population is totally susceptible. 
Matrices $F$ and $V$ are
\begin{align}
\label{FV}
F =& \beta/N_0 \left[ \begin {array}{cccc} 
S\_u^* & (1-\theta\_w)\,S\_u^* & (1-\theta\_w)\,S\_u^* & (1-\theta\_c)\,S\_u^*\\
S\_n^* & (1-\theta\_w)\,S\_n^* & (1-\theta\_w)\,S\_n^* & (1-\theta\_c)\,S\_n^* \\ 
0&0&0&0\\
0&0&0&0
 \end {array} \right] \\
 =&\beta/N_0 \left[ \begin {array}{c} S\_u^* \\ S\_n^* \\ 0 \\ 0 \end {array} \right]
        \left[ \begin {array}{cccc} 1, \, 1-\theta\_w, \, 1-\theta\_w, \, 1-\theta\_c \end {array} \right], \text{and}\\  V=&
 \left[ \begin {array}{cccc}  
\hat F_I+\gamma&-\omega&0&0\\
-(1-p_I)\hat F_I&\omega+\gamma&0&0\\
-p_I \hat F_I&0&\omega+\gamma&0\\
0&0&-\omega&\gamma
\end {array} \right].
\end{align}
The matrix inverse of $V$ is 
\begin{align}
\label{vinv}
V^{-1} =&
\frac{1}{\gamma C}
\left[ \begin {array}{cccc}
\gamma(\omega+\gamma)^2 & \gamma\omega(\omega+\gamma)&0&0\\ \noalign{\medskip}
\gamma(\omega+\gamma)(1-p_I) \hat F_I & \gamma(\omega+\gamma)(\hat F_I+\gamma)&0&0 \\ \noalign{\medskip}
\gamma(\omega+\gamma) p_I \hat F_I & \gamma\omega p_I \hat F_I & C \gamma/(\omega+\gamma) & 0 \\ \noalign{\medskip}
\omega(\omega+\gamma) p_I \hat F_I & \omega^2 p_I \hat F_I & C \omega/(\omega+\gamma) & C
\end {array} \right],
\end{align}
where $C= \big( \gamma(\omega+\gamma)+(\gamma+\omega p_I)\hat F_I \big) (\omega+\gamma)$ and $\hat F_I$ is the \percap testing rate for the infected people and represented in Eq. \eqref{eq:fi}. Note that all the columns of matrix $V^{-1}$ summ up to $1/\gamma$.

The particular form of $F$ with two rows of zeros at the bottom results in the following blocked form of matrix $G$.
\begin{equation}
\label{mat:G}
G = \left[ \begin {array}{cc}
G_{11}&G_{12}\\
0&0
\end {array} \right],
\end{equation}
where both blocked matricies $G_{11}$ and $G_{12}$ are 2 by 2. Given the upper triangular form of matrix $G$, the basic reproduction number $\Rnum$ (defined as the spectral radius of matrix $G$) is only determined by the blocked matrix $G_{11}$,
\begin{equation}
\label{G11}
G_{11} = \frac{\beta}{\gamma \, C} 
\left[ \begin {array}{c} (\omega-\rho)/\omega \\ \rho/\omega \end {array} \right]
\left[ \begin {array}{cccc} 1, \, 1-\theta\_w, \, 1-\theta\_w, \, 1-\theta\_c \end {array} \right]
\left[ \begin {array}{cc}
\gamma(\omega+\gamma)^2 & \gamma\omega(\omega+\gamma)\\ \noalign{\medskip}
\gamma(\omega+\gamma)(1-p_I)\hat F_I & \gamma(\omega+\gamma)(\hat F_I+\gamma) \\ \noalign{\medskip}
\gamma(\omega+\gamma) p_I \hat F_I & \gamma\omega p_I \hat F_I \\ \noalign{\medskip}
\omega(\omega+\gamma) p_I \hat F_I & \omega^2 p_I \hat F_I
\end {array} \right].
\end{equation}
It is notable that matrix $F$ \eqref{FV} has rank one and consequently $G_{11}$ does so. That is $G_{11}$ has only one non-zero eigenvalue which is $\Rnum$.

@@

This is to provide a mathematical materials to prove that $\pro \Delta$ can be positive or negative, with $\Delta$ defined in Eq. \eqref{eq:abc}, and thus $\pro \Rnum < 0$, where $\Rnum$ is the basic reproduction number, given in Eqs.\,\eqref{R0}. We can rewrite matrix $G_{11}$ in \eqref{G11} in the following form to simplify the calculations.
\begin{equation}
\label{G112}
G_{11} = \frac{\beta}{\gamma \, C} 
\left[ \begin {array}{c}  (\omega-\rho)/\omega \\ \rho/\omega  \end {array} \right]
\left[ \begin {array}{cccc} 
C-C_1, C-C_1-C_2\end {array} \right],
\end{equation}
where $C$ is the same as the one in Eq. \eqref{eq:abc}, i.e.,
$$C=(\omega+\gamma)(\gamma(\omega+\gamma)+(\omega\,p_I+\gamma)\,\hat F_I),$$
and $C_1$ and $C_2$ are 
\begin{align}
\label{eq:C12}
C_1 =& (\omega+\gamma)(\theta\_w \, \gamma+\theta\_c\, \omega\, p_I) \hat F_I,\\
C_2 =& (\omega+\gamma)\gamma^2\,\theta\_w+(\theta\_w-\theta\_c)\gamma\,\omega\,p_I\,\hat F_I,
\end{align}
where $\hat F_I$ is given in Eq. \eqref{eq:fi}.
$\Rnum$ is in the same form as in Eq. \eqref{R0}  
$$\Rnum= \frac{\beta}{\gamma} (1-\Delta).$$

>>>>

\begin{figure}[h!]
\centering
\begin{subfigure}[t]{.45\textwidth}
\centering
\includegraphics[width=\linewidth]{codes/R0contour_random2.pdf}
\caption{}
\end{subfigure}
%
\begin{subfigure}[t]{.45\textwidth}
\centering
\includegraphics[width=\linewidth]{codes/R0contour_TTI2.pdf}
\caption{}
\end{subfigure}
\caption{
{\bf A comparison of the behaviour of the basic reproduction number, $\Rnum$, between random versus targeted testing strategies at different levels of testing and isolation.}
We numerically evaluate $\Delta$ \eqref{R0}, reflecting the reduction of $\Rnum$ with respect to testing and isolation. We use the following parameters (listed in Table \ref{tab:params}):
$N_0=1 \times 10^6$, $\omega \in (0.5,2]$ 1/day, $1/\gamma= 3$ days, $\rho \in [0,0.5]$ 1/day, $\theta\_w$ and $\theta\_c$ vary between 0 and 1 with 0 for no effect and 1 for full effect of isolation on the transmission probability, $\beta=0.39$ 1/day, $p_S=0$, $p_I=1$ and $p_R=0.5$. Contours of $\Delta$ are plotted for two testing strategies identified by a set of relative testing weights; (a) random testing where $w_S=w_I=w_R=1$ and (b) targeted testing where $w_S=0.3$ and $w_I=w_R=1$. The black solid line in each panel represents the critical contour of $\Delta=1-\frac{\gamma}{\beta}$, i.e., the $\Delta$ corresponding to the threshold of $\Rnum=1$. \bmb{we still have some collisions in the y-axis tick labels}
}
\label{pan2}
\end{figure}

@@

The model in the form of a system of ordinary differential equations is 
\begin{subequations}\label{model}
\begin{align}
 d S\_u/dt &= -\Lambda S\_u - F_S S\_u + \omega S\_n, \\
 d S\_n/dt &= -(1-\theta\_w)\Lambda S\_n + (1-p_S) F_S S\_u - \omega S\_n, \\
 d S\_p/dt &= -(1-\theta\_w)\Lambda S\_p + p_S F_S S\_u - \omega S\_p, \\
 d S\_c/dt &= -(1-\theta\_c)\Lambda S\_c + \omega S\_p, \\
 d I\_u/dt &= \Lambda S\_u - F_I I\_u + \omega I\_n  - \gamma I\_u, \\
 d I\_n/dt &= (1-\theta\_w)\Lambda S\_n + (1-p_I) F_I I\_u - \omega I\_n -\gamma I\_n, \\
 d I\_p/dt &= (1-\theta\_w)\Lambda S\_p + p_I F_I I\_u - \omega I\_p -\gamma I\_p, \\
 d I\_c/dt &= (1-\theta\_c)\Lambda S\_c + \omega I\_p - \gamma I\_c, \\
 d R\_u/dt &= \gamma I\_u - F_R R\_u + \omega R\_n, \\
 d R\_n/dt &= \gamma I\_n + (1-p_R) F_R R\_u - \omega R\_n, \\
 d R\_p/dt &= \gamma I\_p + p_R F_R R\_u  - \omega R\_p, \\
 d R\_c/dt&= \gamma I\_c + \omega R\_p, \\
 dN/dt &= \omega (S\_n + I\_n + R\_n),  \\
 dP/dt &= \omega(I\_p + R\_p) ,
\end{align}
\end{subequations}

where parameters are specified in Table \ref{tab:params}. The next generation matrix for this model is $G = F V^{-1}$, where matrix $F$ represents the inflow of new infection to the infected compartments and matrix $V$ represents the flow in the infected compartments when the population is totally susceptible. 
Matrices $F$ and $V$ are
\begin{align}
\label{FV}
F =& \beta/N_0 \left[ \begin {array}{cccc} 
S\_u^* & (1-\theta\_w)\,S\_u^* & (1-\theta\_w)\,S\_u^* & (1-\theta\_c)\,S\_u^*\\
(1-\theta\_w)S\_n^* & (1-\theta\_w)^2\,S\_n^* & (1-\theta\_w)^2\,S\_n^* & (1-\theta\_w)(1-\theta\_c)\,S\_n^* \\ 
0&0&0&0\\
0&0&0&0
 \end {array} \right] \\
 =&\beta/N_0 \left[ \begin {array}{c} S\_u^* \\ (1-\theta\_w)S\_n^* \\ 0 \\ 0 \end {array} \right]
        \left[ \begin {array}{cccc} 1, \, 1-\theta\_w, \, 1-\theta\_w, \, 1-\theta\_c \end {array} \right], \text{and}\\  V=&
 \left[ \begin {array}{cccc}  
\hat F_I+\gamma&-\omega&0&0\\
-(1-p_I)\hat F_I&\omega+\gamma&0&0\\
-p_I \hat F_I&0&\omega+\gamma&0\\
0&0&-\omega&\gamma
\end {array} \right].
\end{align}
The matrix inverse of $V$ is 
\begin{align}
\label{vinv}
V^{-1} =&
\frac{1}{\gamma C}
\left[ \begin {array}{cccc}
\gamma(\omega+\gamma)^2 & \gamma\omega(\omega+\gamma)&0&0\\ \noalign{\medskip}
\gamma(\omega+\gamma)(1-p_I) \hat F_I & \gamma(\omega+\gamma)(\hat F_I+\gamma)&0&0 \\ \noalign{\medskip}
\gamma(\omega+\gamma) p_I \hat F_I & \gamma\omega p_I \hat F_I & C \gamma/(\omega+\gamma) & 0 \\ \noalign{\medskip}
\omega(\omega+\gamma) p_I \hat F_I & \omega^2 p_I \hat F_I & C \omega/(\omega+\gamma) & C
\end {array} \right],
\end{align}
where $C= \big( \gamma(\omega+\gamma)+(\gamma+\omega p_I)\hat F_I \big) (\omega+\gamma)$ and $\hat F_I$ is the \percap testing rate for the infected people and represented in Eq. \eqref{eq:fi}. Note that all the columns of matrix $V^{-1}$ summ up to $1/\gamma$.

The particular form of $F$ with two rows of zeros at the bottom results in the following blocked form of matrix $G$.
\begin{equation}
\label{mat:G}
G = \left[ \begin {array}{cc}
G_{11}&G_{12}\\
0&0
\end {array} \right],
\end{equation}
where both blocked matricies $G_{11}$ and $G_{12}$ are 2 by 2. Given the upper triangular form of matrix $G$, the basic reproduction number $\Rnum$ (defined as the spectral radius of matrix $G$) is only determined by the blocked matrix $G_{11}$,
\begin{equation}
\label{G11}
G_{11} = \frac{\beta}{\gamma \, C} 
\left[ \begin {array}{c} (\omega-\rho)/\omega \\ (1-\theta\_w)\rho/\omega \end {array} \right]
\left[ \begin {array}{cccc} 1, \, 1-\theta\_w, \, 1-\theta\_w, \, 1-\theta\_c \end {array} \right]
\left[ \begin {array}{cc}
\gamma(\omega+\gamma)^2 & \gamma\omega(\omega+\gamma)\\ \noalign{\medskip}
\gamma(\omega+\gamma)(1-p_I)\hat F_I & \gamma(\omega+\gamma)(\hat F_I+\gamma) \\ \noalign{\medskip}
\gamma(\omega+\gamma) p_I \hat F_I & \gamma\omega p_I \hat F_I \\ \noalign{\medskip}
\omega(\omega+\gamma) p_I \hat F_I & \omega^2 p_I \hat F_I
\end {array} \right].
\end{equation}
It is notable that matrix $F$ \eqref{FV} has rank one and consequently $G_{11}$ does so. That is $G_{11}$ has only one non-zero eigenvalue which is $\Rnum$.

@@

This is to provide a mathematical materials to prove that $\pro \Delta$ can be positive or negative, with $\Delta$ defined in Eq. \eqref{eq:abc}, and thus $\pro \Rnum < 0$, where $\Rnum$ is the basic reproduction number, given in Eqs.\,\eqref{R0}. We can rewrite matrix $G_{11}$ in \eqref{G11} in the following form to simplify the calculations.
\begin{equation}
\label{G112}
G_{11} = \frac{\beta}{\gamma \, C} 
\left[ \begin {array}{c}  (\omega-\rho)/\omega \\ (1-\theta\_w)\rho/\omega  \end {array} \right]
\left[ \begin {array}{cccc} 
C-C_1, C-C_1-C_2\end {array} \right],
\end{equation}
where $C$ is the same as the one in Eq. \eqref{eq:abc}, i.e.,
$$C=(\omega+\gamma)(\gamma(\omega+\gamma)+(\omega\,p_I+\gamma)\,\hat F_I),$$
and $C_1$ and $C_2$ are 
\begin{align}
\label{eq:C12}
C_1 =& (\omega+\gamma)(\theta\_w \, \gamma+\theta\_c\, \omega\, p_I) \hat F_I,\\
C_2 =& (\omega+\gamma)\gamma^2\,\theta\_w+(\theta\_w-\theta\_c)\gamma\,\omega\,p_I\,\hat F_I,
\end{align}
where $\hat F_I$ is given in Eq. \eqref{eq:fi}.
$\Rnum$ is in the same form as in Eq. \eqref{R0}  
$$\Rnum= \frac{\beta}{\gamma} (1-\Delta).$$
